Supplementary Materials: Separating random and deterministic sources of computational noises in explore-exploit decisions

### 2025-04-19 Plos Comp Bio - revision v1/manuscript_DeterministicRandomNoise.tex
reorganize codes

% control the fontsize
\PassOptionsToPackage{textsize=scriptsize}{todonotes}
%\usepackage[final]{changes}
\usepackage{changes}
\definechangesauthor[name={Siyu Wang}, color=red]{siyu}
\definechangesauthor[name={Bob}, color=blue]{bob}

@@

	\newpage
	\begin{abstract}
		Human decision making is inherently variable. While this variability is often seen as a sign of suboptimal behavior, recent work suggests that variability can actually be adaptive. An example arises when we must choose between exploring unknown options or exploiting options we know well. A little randomness in these `explore-exploit' decisions is remarkably effective as it can encourage us to explore options we might otherwise ignore. In line with this idea, several studies have found evidence that people increase their behavioral variability when it is valuable to explore. A key question, however, is whether this variability in so-called `random exploration' is actually random. That is, is random exploration driven by stochastic processes in the brain or by some unobserved deterministic process that we have failed to account for when measuring behavioral variability? By designing an explore-exploit task in which, unbeknownst to them, participants are presented with the exact same choice twice, we provide a partial answer to this question. By modeling behavior in this task, we were able to estimate a lower bound on the amount of variability that is deterministically driven by the stimulus and an upper bound on the amount of variability that is random. Using this approach, we find evidence that at least 14$\%$ of the variability in random exploration in our studied task can be accounted for by deterministic processing of the stimulus. Conversely, this suggests that up to 86$\%$ of the variability is truly `random', although it is still possible that this variability is driven by deterministic factors not related to the stimulus. Finally, our results suggest that both deterministic and random sources of variability change proportionally to each other as the value of exploration increases, suggesting that a common noise gating mechanism may be at play in random exploration.		
	\end{abstract}
	\newpage
	\section*{Introduction}
	%BOB SAID: IN GENERAL, NEED TO GO THROUGH THE PAPER AND MAKE SURE THAT `RANDOM' NOISE IS REPLACES WITH `NOT A DETERMINISTIC FUNCTION OF THE STIMULUS' OR SOMETHING LIKE THAT

@@

	It has recently been shown that humans appear to use random exploration and can increase  decision noise when it is more beneficial to explore \citep{Gershman2018, wilson2014, Findling19}. In one of these tasks, known as the Horizon Task \citep{wilson2014}, the key manipulation is the horizon condition, i.e. the number of decisions remaining for the participant to make. Increasing the horizon makes exploration more valuable as there is more time to use the information gained by exploration to maximize future rewards. For example, if you are leaving town tomorrow (short horizon), you will probably exploit the restaurant you know and love, but if you are in town for a while (long horizon), you will be more likely to explore the new restaurant. Using such a horizon manipulation it has been shown that people's behavior is more variable in long horizons than short horizons, suggesting that they use adaptive decision noise to solve the explore-exploit dilemma \citep{wilson2014}.

	One limitation of this previous research, however, is that it is difficult to tell whether what we have called `decision noise' actually reflect a noise process. From a modeling perspective, decision noise as defined in previous research essentially quantifies the extent to which behavior  cannot be explained by a computational model. A missing deterministic component from the model could give rise to variability in behavior that might appear to be random noise. For example, in the restaurant example, my usual preference for one restaurant or another may be overruled if I see an ex romantic partner going into one of them. Avoiding an ex is a deterministic process, but if we fail to take the ex's presence into account as scientists modeling the decision, then over a series of such decisions where the ex is present or not, we would mistakenly attribute the ensuing `variability' in choice to randomness.

@@

	\subsection*{Both behavioral variability and information seeking increase with horizon}
	Before discussing the results for repeated games, we first confirm that the basic behavior in this task is consistent with our previously reported results using both a model-free and model-based approach \citep{wilson2014}. In both analyses, we focus on just the first free-choice trial in each game, where the only thing that differs between the horizon conditions is the number of choices that participants will make in the future.

@@

	Here, $n$ denotes decision noise, which, in this version of the model is a combination of deterministic and random noise. $n$ is assumed to come from a logistic distributions with mean 0 and standard deviations $\sigma$.

@@

As shown in Figure \ref{fig:mf2}, people's behavior falls in between the pure deterministic noise prediction and the pure random noise prediction. Specifically, behavior is different from the pure random noise prediction in the both the [1 3] condition (t(64) = 4.83, p $<$ 0.001 for horizon 1, t(64) = 3.12 p $=$ 0.003 for horizon 6) and the [2 2] condition (t(64) = 3.92, p $<$ 0.001 for horizon 1, t(64) = 3.71, p $<$ 0.001 for horizon 6). Likewise, behavior is different from pure deterministic noise prediction in both the [1 3] condition (t(64) = 13.72, p $<$ 0.001 for horizon 1, t(64) = 16.71, p $<$ 0.001 for horizon 6) and the [2 2] condition (t(64) = 9.55, p $<$ 0.001 for horizon 1, t(64) = 17.93, p $<$ 0.001 for horizon 6). As a negative control of our method for estimating $p(\mbox{inconsistent})$ for purely random noise, we simulated choices using a decision model that only includes random noise (Equation \ref{eq:newmodel}), and found that $p(\mbox{inconsistent})$ in this simulated data is not different from our pure random noise prediction in all horizon and uncertainty conditions (p $>$ 0.05, Supplementary Figure S3). Together, our results suggest that both random noise and deterministic noise contribute to the choice variability in random exploration. However, the relative contribution from each of these types of noise, as well as how each type of noise changes with horizon, are difficult to discern.

@@

\subsubsection*{Posterior predictive checks}
In addition to fitting the model to behavior, it is also important to check whether the model captures the qualitative patterns of the data \citep{Wilson2019} --- specifically how p(high info), p(low mean) and p(inconsistent) change with horizon.

@@

Although the random noise interpretation is theoretically appealing, our approach, while an improvement on previous methods, is not without limitations. Most important is that our measure of `random' noise is only an upper bound on the true level of randomness and that, in principle, the random decision noise could be lower. Specifically, in our model, what we labeled random noise was really `non-stimulus-driven variability'. While this non-stimulus-driven variability could be driven by truly random stochastic processes, it could also be driven by deterministic processing that is unrelated to the stimuli in the task. For example, such deterministic noise could be driven by differences in where people look, or for how long they look, or by whether they were fidgeting or scratching their nose \citep{Musall2019}. In addition to this conceptual limitation in measuring deterministic noise, parameter recovery simulations suggest that our estimation method also slightly underestimates deterministic noise (see Figure \ref{fig:paramrecover_main}, Supplementary Figure S5, S6). As a result, from both a conceptual and methodological perspective, it is possible that the remaining 86\% of the decision noise that is not stimulus-driven noise, could be deterministic.

Like the random noise account, the deterministic noise account is also in line with previous work in which neural variability can be accounted for by fluctuations in sensory inputs. For example, MT neurons were shown to have a reproducible temporal modulation in response to a fixed random motion stimuli \citep{Bair1996}. In other words, `irrelevant' features in the stimuli are represented in a reliable way in the brain that could drive downstream choices in a predictable way.

@@

To see how a change in reward processing could affect random and deterministic noise, consider the simple decision model we introduced in Equation \ref{eq:newmodel}. In this model, choice is determined by the sign of the difference in utility $\Delta Q$ between the two options, where
\begin{equation}
	\label{eq:noise1}
	\Delta Q = 
	\Delta R + A \Delta I + b + n_{det} + n_{ran}
\end{equation}
Now imagine a case where the reward signal is scaled by a factor $\beta$.  In this case, $\Delta Q$ becomes
\begin{equation}
	\label{eq:noise2}
	\Delta Q = 
	\beta \Delta R + A \Delta I + b + n_{det} + n_{ran}
\end{equation}
Because the choice only depends on the sign of $\Delta Q$, scaling $\Delta Q$ by a factor of $1/\beta$ will not change the behavior of the model.  Thus, if we divide both sides of the above equation by $\beta$ we get
\begin{equation}
	\label{eq:noise3}
	\Delta Q / \beta = 
	\Delta R + A \Delta I / \beta + b / \beta + n_{det} / \beta + n_{ran} / \beta
\end{equation}
which is equivalent to a scaling of both deterministic and random noise by the same factor $1/\beta$. Thus, one interpretation of our result that both deterministic and random noise change across horizons with the same ratio, is that this reflects a change in reward processing. That is, the reward signal is reduced in the longer horizon condition (smaller $\beta$ in horizon 6 than horizon 1).  %In the alternate interpretation, where deterministic and random noise are varied independently it is not clear why they should change by the same scale factor.

Such a reduction in the strength of reward coding in exploration, is consistent with our recent work using a drift diffusion model (DDM) to model explore-exploit decisions \citep{Feng21}. In the drift diffusion model, changes in behavioral variability can be driven by changes in the decision threshold (smaller threshold = more noise) or changes in the signal-to-noise ratio with which reward is encoded (lower SNR = more noise). By fitting both choices and response times, we were able to distinguish between these two accounts showing that the majority of the horizon-change in variability was driven by changes in SNR not threshold. However, this model could not determine whether the changes in SNR were driven by signal or noise. By showing that the change in deterministic and random noise have approximately the same ratio, the present work suggests that this SNR change is driven by changes in reward-signal processing, not noise. Of course, to truly see whether changes in signal or noise are driving random exploration will require more direct measurements of neural processing such as with neuroimaging and electrophysiology \citep{Tomov20, Ebitz18,costa1,costa22}

>>>>

% control the fontsize
\PassOptionsToPackage{textsize=scriptsize}{todonotes}
\usepackage[]{changes}
\usepackage{changes}
\definechangesauthor[name={Siyu Wang}, color=red]{siyu}
\definechangesauthor[name={Bob}, color=blue]{bob}

@@

	\newpage
	\begin{abstract}
		Human decision making is inherently variable. While this variability is often seen as a sign of suboptimal behavior, \added[id=siyu]{both theoretical work in machine learning and empirical human studies} suggest that variability can actually be adaptive. An example arises when we must choose between exploring unknown options or exploiting options we know well. A little randomness in these `explore-exploit' decisions is remarkably effective as it can encourage us to explore options we might otherwise ignore. In line with this idea, several studies have found evidence that people increase their behavioral variability when it is valuable to explore. A key question, however, is whether this variability in so-called `random exploration' is actually random. That is, is random exploration driven by stochastic processes in the brain or by some unobserved deterministic process that we have failed to account for when measuring behavioral variability? By designing an explore-exploit task in which, unbeknownst to them, participants are presented with the exact same choice twice, we provide a partial answer to this question. By modeling behavior in this task, we were able to estimate a lower bound on the amount of variability that is deterministically driven by the stimulus and an upper bound on the amount of variability that is random. Using this approach, we find evidence that at least 14$\%$ of the variability in random exploration in our studied task can be accounted for by deterministic processing of the stimulus. Conversely, this suggests that up to 86$\%$ of the variability is truly `random', although it is still possible that this variability is driven by deterministic factors not related to the stimulus. Finally, our results suggest that both deterministic and random sources of variability change proportionally to each other as the value of exploration increases, suggesting that a common noise gating mechanism may be at play in random exploration.		
	\end{abstract}
	\newpage
	\section*{Introduction}
	%BOB SAID: IN GENERAL, NEED TO GO THROUGH THE PAPER AND MAKE SURE THAT `RANDOM' NOISE IS REPLACES WITH `NOT A DETERMINISTIC FUNCTION OF THE STIMULUS' OR SOMETHING LIKE THAT

@@

	It has recently been shown that humans appear to use random exploration and can increase decision noise when it is more beneficial to explore \citep{Gershman2018, wilson2014}, \added[id=siyu]{as has also been suggested in computational models of animal behavior \citep{DOYA2002495, KHAMASSI2013441}}. In one of these tasks, known as the Horizon Task \citep{wilson2014}, the key manipulation is the horizon condition, i.e. the number of decisions remaining for the participant to make. Increasing the horizon makes exploration more valuable as there is more time to use the information gained by exploration to maximize future rewards. For example, if you are leaving town tomorrow (short horizon), you will probably exploit the restaurant you know and love, but if you are in town for a while (long horizon), you will be more likely to explore the new restaurant. Using such a horizon manipulation it has been shown that people's behavior is more variable in long horizons than short horizons, suggesting that they use adaptive decision noise to solve the explore-exploit dilemma \citep{wilson2014}.

	One limitation of this previous research, however, is that it is difficult to tell whether what we have called `decision noise' actually reflect a noise process. From a modeling perspective, decision noise as defined in previous research essentially quantifies the extent to which behavior cannot be explained by a computational model. A missing deterministic component from the model could give rise to variability in behavior that might appear to be random noise. For example, in the restaurant example, my usual preference for one restaurant or another may be overruled if I see an ex romantic partner going into one of them. Avoiding an ex is a deterministic process, but if we fail to take the ex's presence into account as scientists modeling the decision, then over a series of such decisions where the ex is present or not, we would mistakenly attribute the ensuing `variability' in choice to randomness. \added[id=siyu]{To dissociate a missing deterministic component from a true random process, choice consistency between repeated decisions can be utilized to decompose behavioral variability into predictable deterministic components and unpredictable random components \citep{WYART2016, Wyart_2018, Findling19, Findling21}.}

@@

	\subsection*{Both behavioral variability and information seeking increase with horizon}
	Before discussing the results for repeated games, we first confirm that the basic behavior in this task is consistent with our previously reported results using both a model-free and model-based approach \citep{wilson2014}. In both analyses, we focus on just the first free-choice trial in each game, where the only thing that differs between the horizon conditions is the number of choices that participants will make in the future. \added[id=siyu]{Subsequent choices in Horizon 6 games were not analyzed.}

@@

	Here, $n$ denotes decision noise, which, in this version of the model is a combination of deterministic and random noise. $n$ is assumed to come from a logistic distribution with mean 0 and standard deviations $\sigma$.

@@

As shown in Figure \ref{fig:mf2}, people's behavior falls in between the pure deterministic noise prediction and the pure random noise prediction. Specifically, behavior is different from the pure random noise prediction in both the [1 3] condition (t(64) = 4.83, p $<$ 0.001 for horizon 1, t(64) = 3.12 p $=$ 0.003 for horizon 6) and the [2 2] condition (t(64) = 3.92, p $<$ 0.001 for horizon 1, t(64) = 3.71, p $<$ 0.001 for horizon 6). Likewise, behavior is different from pure deterministic noise prediction in both the [1 3] condition (t(64) = 13.72, p $<$ 0.001 for horizon 1, t(64) = 16.71, p $<$ 0.001 for horizon 6) and the [2 2] condition (t(64) = 9.55, p $<$ 0.001 for horizon 1, t(64) = 17.93, p $<$ 0.001 for horizon 6). As a negative control of our method for estimating $p(\mbox{inconsistent})$ for purely random noise, we simulated choices using a decision model that only includes random noise (Equation \ref{eq:newmodel}), and found that $p(\mbox{inconsistent})$ in this simulated data is not different from our pure random noise prediction in all horizon and uncertainty conditions (p $>$ 0.05, Supplementary Figure S3). Together, our results suggest that both random noise and deterministic noise contribute to the choice variability in random exploration. However, the relative contribution from each of these types of noise, as well as how each type of noise changes with horizon, are difficult to discern.

@@

\subsubsection*{Posterior predictive checks}
In addition to fitting the model to behavior, it is also important to check whether the model captures the qualitative patterns of the data \citep{Wilson2019, Palminteri} --- specifically how p(high info), p(low mean) and p(inconsistent) change with horizon.

@@

Although the random noise interpretation is theoretically appealing, our approach, while an improvement on previous methods, is not without limitations. Most important is that our measure of `random' noise is only an upper bound on the true level of randomness and that, in principle, the random decision noise could be lower. Specifically, in our model, what we labeled random noise was really `non-stimulus-driven variability'. While this non-stimulus-driven variability could be driven by truly random stochastic processes, it could also be driven by deterministic processing that is unrelated to the stimuli in the task. For example, such deterministic noise could be driven by differences in where people look, or for how long they look, or by whether they were fidgeting or scratching their nose \citep{Musall2019}. \added[id=siyu]{Cross-game deterministic strategies and stimuli from previous games (e.g., memory of previously seen game) were also treated as random noise in our model \citep{WYART2016, Wyart_2018}.} In addition to this conceptual limitation in measuring deterministic noise, parameter recovery simulations suggest that our estimation method also slightly underestimates deterministic noise (see Figure \ref{fig:paramrecover_main}, Supplementary Figure S5, S6). As a result, from both a conceptual and methodological perspective, it is possible that the remaining 86\% of the decision noise that is not stimulus-driven noise, could be deterministic.

Like the random noise account, the deterministic noise account is also in line with previous work in which neural variability can be accounted for by fluctuations in sensory inputs. For example, MT neurons were shown to have a reproducible temporal modulation in response to a fixed random motion stimulus \citep{Bair1996}. In other words, `irrelevant' features in the stimuli are represented in a reliable way in the brain that could drive downstream choices in a predictable way.

@@

\added[id=siyu]{We will show that changes in reward processing can be equivalent to changes in noise in decision models (see also \cite{Cinotti19}).}. To show how a change in reward processing could affect random and deterministic noise, consider the simple decision model we introduced in Equation \ref{eq:newmodel}. In this model, choice is determined by the sign of the difference in utility $\Delta Q$ between the two options, where
\begin{equation}
	\label{eq:noise1}
	\Delta Q = 
	\Delta R + A \Delta I + b + n_{det} + n_{ran}
\end{equation}
Now imagine a case where the reward signal is scaled by a factor $\beta$.  In this case, $\Delta Q$ becomes
\begin{equation}
	\label{eq:noise2}
	\Delta Q = 
	\beta \Delta R + A \Delta I + b + n_{det} + n_{ran}
\end{equation}
Because the choice only depends on the sign of $\Delta Q$, scaling $\Delta Q$ by a factor of $1/\beta$ will not change the behavior of the model.  Thus, if we divide both sides of the above equation by $\beta$ we get
\begin{equation}
	\label{eq:noise3}
	\Delta Q / \beta = 
	\Delta R + A \Delta I / \beta + b / \beta + n_{det} / \beta + n_{ran} / \beta
\end{equation}
which is equivalent to a scaling of both deterministic and random noise by the same factor $1/\beta$. Thus, one interpretation of our result that both deterministic and random noise change across horizons with the same ratio, is that this reflects a change in reward processing. That is, the reward signal is reduced in the longer horizon condition (smaller $\beta$ in horizon 6 than horizon 1).  %In the alternate interpretation, where deterministic and random noise are varied independently it is not clear why they should change by the same scale factor.

Such a reduction in the strength of reward coding in exploration, is consistent with our recent work using a drift diffusion model (DDM) to model explore-exploit decisions \citep{Feng21}. In the drift diffusion model, changes in behavioral variability can be driven by changes in the decision threshold (smaller threshold = more noise) or changes in the signal-to-noise ratio with which reward is encoded (lower SNR = more noise). By fitting both choices and response times, we were able to distinguish between these two accounts showing that the majority of the horizon-change in variability was driven by changes in SNR and not threshold. However, this model could not determine whether the changes in SNR were driven by signal or noise. By showing that the change in deterministic and random noise have approximately the same ratio, the present work suggests that this SNR change is driven by changes in reward-signal processing, not noise. Of course, to truly see whether changes in signal or noise are driving random exploration will require more direct measurements of neural processing such as with neuroimaging and electrophysiology \citep{Tomov20, Ebitz18,costa1,costa22}

### 2025-04-19 Plos Comp Bio - revision v1/supplemental/DetRanNoise_Supplementary.tex
reorganize codes

	\newpage
	\subsection{Parameter recovery of the subject-level parameter fits}
	Next, we tested the ability of our model fitting procedure to recovery parameters from simulated data at the subject level (Figure \ref{fig:paramrecover_session}, \ref{fig:paramrecover_sub}). The correlations between the true vs fitted parameters are significant across participants for all parameters (p $<$ 0.001). The strength of correlation between simulated and fit values are strong for both deterministic noise (R $>$ 0.8, Figure \ref{fig:paramrecover_sub}) and random noise (R $>$ 0.9, Figure \ref{fig:paramrecover_sub}). Despite the strong inter-subject correlations, we again observed a systematic underestimation of $\sigma_{det}$ (Figure \ref{fig:paramrecover_session}, \ref{fig:paramrecover_sub}).

>>>>

	\newpage
	\subsection{Parameter recovery of the subject-level parameter fits}
	Next, we tested the ability of our model fitting procedure to recover parameters from simulated data at the subject level (Figure \ref{fig:paramrecover_session}, \ref{fig:paramrecover_sub}). The correlations between the true vs fitted parameters are significant across participants for all parameters (p $<$ 0.001). The strength of correlation between simulated and fit values are strong for both deterministic noise (R $>$ 0.8, Figure \ref{fig:paramrecover_sub}) and random noise (R $>$ 0.9, Figure \ref{fig:paramrecover_sub}). Despite the strong inter-subject correlations, we again observed a systematic underestimation of $\sigma_{det}$ (Figure \ref{fig:paramrecover_session}, \ref{fig:paramrecover_sub}).
